---
title: "DESeq2_SSW_round2.1"
author: "R Kirsten Tyler"
date: "2/27/2017"
output: html_document
---

###Modeling our data###

Previously, Melissa did this: 

Update on Making and Mapping to a new reference transcriptome:
 - Concatenated fastq files from 4 individuals (over 200M reads) both sick and healthy.
 - Digital normalization by kmer to make a smaller file size (8.7% of original size).
 - Ran Trinity with really low max-memory and CPUs so that it would complete without crashing
 - Predicted ORFs with TransDecoder.pl script.
 - Evaluated contig stats to find that they were fewer, longer contigs.
 - Mapped reads to this new assembly.
 - Extracted read counts
 - Assembled read counts table.
 
#Today!#

1. Move new data
2. DESeq 2
  - 5 models
3. Your models

Our new assembly:
Melissa used the longest isoform
But longest isoform does not equal best isoform
However, by taking the longest one, we are recovering more of our reads

We are using [this tutorial] (https://www.bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2_pdf.pdf)

```{r}
setwd("~/Documents/Ecological_Genomics_PBIO381/SeaStar_Analysis")

library(DESeq2)
library(ggplot2)


countsTable <- read.delim('countsdata_trim2.txt', header = TRUE, stringsAsFactors = TRUE, row.names = 1)
countData <- as.matrix(countsTable)
head(countData)
```


```{r}
conds <- read.delim("cols_data_trim.txt", header = TRUE, stringsAsFactors = TRUE, row.names = 1)
head(conds)
```


```{r}
colData <- as.data.frame(conds)
head(colData)
tail(colData)
```

# MODEL NUMBER 1: TEST EFFECT OF HEALTH CONTROLLING FOR LOCATION #
```{r}
dds1 <- DESeqDataSetFromMatrix(countData = countData, colData = colData, design = ~ location + health) 
```

# MODEL H vs. S within intertidal
# MODEL H vs. S within subtidal 
```{r}
ddsI <- DESeqDataSetFromMatrix(countData = countData, colData = colData, design = ~ location$int + health)
ddsS <- DESeqDataSetFromMatrix(countData = countData, colData = colData, design = ~ location$sub + health) 
```


```{r}

# Here, we are asking: what's the effect of health given the location?
# so we are focusing in on the LAST term as the one that is the main effect
# in this typical model, the "health effect" represents the overall effect of health status, controlling for differences due to location

dim(dds1)
dds1 <- dds1[ rowSums(counts(dds1)) > 100, ]
dim(dds1)

```



##For practice, let's work with fewer genes so that the models can run faster...##
```{r}
#dds <- dds[sample(nrow(dds), 1200), ] #randomly sample of 1200 rows from our 12000 row data matrix. when we do our real analyses, we won't want to sample - we'll want to use the entire data matrix
#dim(dds)
```


```{r}

colData(dds1)$health <- factor(colData(dds1)$health, levels=c("H","S")) #sets that "healthy" is the reference

dds1 <- DESeq(dds1) #Running the model
```


```{r}
res1 <- results(dds1)
res1 <- res1[order(res1$padj),] #this sorts our results by the adjusted p-value. Nice!
head(res1)
```

```{r}
summary(res1)
```

We can see here that (up = more highly expressed in sick) + (down = more highly expressed in healthy) genes are differentially expressed between the healthy and sick individuals in our data set



# MODEL NUMBER 2 - INTERACTIONS #

```{r}
dds <- DESeqDataSetFromMatrix(countData = countData, colData = colData, design = ~ location + health + location:health) #now we are adding "interaction of location and health"

dim(dds)
```


```{r}
dds <- dds[ rowSums(counts(dds)) > 100, ]

dds <- dds[ sample(nrow(dds), 1200), ]
dim(dds)

colData(dds)$health <- factor(colData(dds)$health, levels=c("H","S"))  #sets that "healthy is the reference

dds <- DESeq(dds, parallel=T) #Running the model (parallel = using two cores on my computer = wicked fast!!)
```


```{r}
resultsNames(dds) #checking out the results headers
```


```{r}
res <- results(dds, name="locationsub.healthS")
res <- res[order(res$padj),] #ordering the results (genes) by adjusted p-value (most significant will be at the top of the results matrix)
head(res) # a list of the genes by significance

```



```{r}
summary(res)
```

# MODEL NUMBER 3 - GROUP DESIGNS can be used for contrasts of interest or interactions #
```{r}
colData$group <- factor(paste0(colData$location, ".", colData$health, ".", colData$score)) #here we are pasting together different variables in which we are interested into a group. This helps us focus on subsets of data. We can focus on specific parts of data: a specific day, for example.
dds <- DESeqDataSetFromMatrix(countData = countData, colData = colData, design = ~ group) 
dim(dds)
head(colData$group)
```


```{r}
dds <- dds[ rowSums(counts(dds)) > 100, ]

dds <- dds[sample(nrow(dds), 1200), ]
dim(dds)

dds <- DESeq(dds, parallel=T)
```


```{r}
resultsNames(dds) #checking out the results 

```


```{r}
res <- results(dds, contrast=list( c("groupint.H.0","groupsub.H.0"), c("groupint.S.1","groupsub.S.1")), listValues=c(1/2, -1/2)) #means have to equal to zero. We are balancing the results

res <- res[order(res$padj),]
head(res)
```

```{r}
summary(res)
```


# MODEL NUMBER 4 - TIME SERIES USING REDUCED MODELS #
## based on example on [bioconductor]: (http://www.bioconductor.org/help/workflows/rnaseqGene/#running-the-differential-expression-pipeline)

```{r}
ddsTS <- DESeqDataSetFromMatrix(countData = countData, colData = colData, ~ health + day + health:day)
ddsTS <- ddsTS[ rowSums(counts(ddsTS)) > 100, ]

ddsTS <- ddsTS[sample(nrow(ddsTS), 1200), ]
dim(ddsTS) 
```


```{r}
ddsTS <- DESeq(ddsTS, parallel=T, test="LRT", reduced = ~ health + day) #log likelihood ratio test = look at distributions to see if there are differences. Does this model significantly show the variance in the data? Which model does it better?

#Questions: what does "reduced" mean? What is the test "LRT"? Take the interaction out and test the likelihood of that model again
resTS <- results(ddsTS)
resTS$symbol <- mcols(ddsTS)$symbol

head(resTS[order(resTS$padj),],4)
```


```{r}
summary(resTS)
```

# MODEL NUMBER 5 - EFFECT OF DISEASE SEVERITY SCORE #

```{r}
dds <- DESeqDataSetFromMatrix(countData = countData, colData = colData, design = ~ score) 
dim(dds)

dds <- dds[ rowSums(counts(dds)) > 100, ]

#PRACTICE SAMPLE (COMMENT OUT THE LINE WHEN PERFORMING REAL ANALYSIS)
dds <- dds[sample(nrow(dds), 1200), ]
dim(dds)
```


```{r}
dds <- DESeq(dds, parallel=T) #Running the model

res <- results(dds)
res <- res[order(res$padj),]
head(res)
```


```{r}
summary(res)
```


# MODEL NUMBER 6 thru INFINITY - YOUR CHOICE...#

#Data quality assessment by sample clustering and visualization#
```{r}
plotMA(res, main="DESeq2", ylim=c(-2,2))
```


```{r}
## Check out one of the genes to see if it's behaving as expected....
d <- plotCounts(dds, gene="TRINITY_DN44744_c1_g2_TRINITY_DN44744_c1_g2_i2_g.17686_m.17686", intgroup=(c("health","day","location")), returnData=TRUE) #plotCounts is pulling out the gene of interest and looking at different factors of it
d
```

```{r}
p <- ggplot(d, aes(x= health, y=count, shape = location)) + theme_minimal() + theme(text = element_text(size=20), panel.grid.major = element_line(colour = "grey"))
p <- p + geom_point(position=position_jitter(w=0.3,h=0), size = 3) + ylim(0,500)
p
```

```{r}
## Check out one of the genes to see interaction between score, health and expression....
d <- plotCounts(dds, gene="TRINITY_DN46245_c3_g3_TRINITY_DN46245_c3_g3_i2_g.21719_m.21719", intgroup=(c("health","score","location")), returnData=TRUE)
d
```


```{r}
p <- ggplot(d, aes(x= score, y=count, shape = health, color = location)) + theme_minimal() + theme(text = element_text(size=20), panel.grid.major = element_line(colour = "grey"))
p <- p + geom_point(position=position_jitter(w=0.3,h=0), size = 3) 
p
```


```{r}
p <- ggplot(d, aes(x=score, y=count, color=health, group=health)) 
p <- p +  geom_point() + stat_smooth(se=FALSE,method="loess") +  scale_y_log10()
p
```


```{r}
############## PCA plots
vsd <- varianceStabilizingTransformation(dds, blind=FALSE)

plotPCA(vsd, intgroup=c("score")) #we are looking at health score (0 = healthy; 5 = dead; 1,2,3,4 = sick to different extents)
```


```{r}
plotPCA(vsd, intgroup=c("health"))
```


```{r}
plotPCA(vsd, intgroup=c("day"))
```


```{r}
plotPCA(vsd, intgroup=c("location"))
```


```{r}
plotPCA(vsd, intgroup=c("health","location"))
```




